A Mechanised Library for Generic (Co)Programming

### 2024-02-29 paper/paper.tex
Centered a bunch of stuff and inserted cit for recursive coalgebras.

For example, assuming the type of lists over a type \coq{A},
namely \coq{List A}, and given an
algebra \coq{alg : unit + A * List A -> List A} a
catamorphism \coq{cata alg : List A -> B} destructs a list into its head and
tail using a destructor \coq{InOp : List A -> unit + A * List A}. This latter is
actually an isomorphim with its inverse \coq{In : unit + A * List A -> List A}
defined by the maps \coq{Empty : unit -> List A}
and \coq{Cons : A * List A -> List A}.

After destructing the input the catamorphism is applying itsefl recursively.
This is done using functoriality of the shape of the data, which is a map of
type \coq{unit + A * List A -> unit + A * List B}. Once the recursion as
returned it result it applies the algebra \coq{alg} which composes together the
results of the recursion yielding the result of type \coq{B}.

More abstractly, assuming a base functor \coq{F} a catamorphism is the unique solution to the equation
\coq{cata alg = alg . fmap (cata alg) . InOp}.

@@

% C
Hylomorphism are more general recursion schemes capturing the
\emph{divide-and-conquer} idea. In short, the input is destructured by its not
structurally. In the example, we gave in the introduction the destructuring
program is given by \coq{divide} which is not an isomorphism.  The question the
hylomorphism satisfies is more general than the catamorphism in that the
destructing map \coq{InOp} is generalised to a \emph{recursive coalgebra}, that
is intuitively a coalgebra that can be applied only a finite amount of times.
In fact, in the case of \coq{divide} this is a recursive coalgebra because the
size of the list is getting smaller.

@@

  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
This definition, however, prevents us from composing catamorphisms and
anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular:
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite
trees, then their return type can also be the \emph{least} fixed point of the
container extension. We encode in Coq the notion of \emph{recursive
anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.  Recursive
coalgebras are coalgebras whose anamorphism produces finite trees.  The
following predicate captures this fact, \coq{h} is recursive if
\coq{h} can only be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times (\coq{RecF}).  Note
that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

>>>>

For example, assuming the type of lists over a type \coq{A},
namely \coq{List A}, and given an
algebra \coq{alg : unit + A * List A -> List A} a
catamorphism \coq{cata alg : List A -> B} destructs a list into its head and
tail using a destructor \coq{InOp : List A -> unit + A * List A}. This latter is
actually an isomorphim with its inverse
\begin{minted}{coq}
In : unit + A * List A -> List A
\end{minted}

defined by the maps \coq{Empty : unit -> List A}
and \coq{Cons : A * List A -> List A}.

After destructing the input the catamorphism is applying itsefl recursively.
This is done using functoriality of the shape of the data, which is a map of
type
\begin{minted}{coq}
unit + A * List A -> unit + A * List B
\end{minted}
Once the recursion as returned it result it applies the algebra \coq{alg} which
composes together the results of the recursion yielding the result of
type \coq{B}.

More abstractly, assuming a base functor \coq{F} a catamorphism is the unique
solution to the equation
\begin{minted}{coq}
cata alg = alg . fmap (cata alg) . InOp
\end{minted}

@@

% C
Hylomorphism are more general recursion schemes capturing the
\emph{divide-and-conquer} idea. In short, the input is destructured by its not
structurally. In the example, we gave in the introduction the destructuring
program is given by \coq{divide} which is not an isomorphism.  The question the
hylomorphism satisfies is more general than the catamorphism in that the
destructing map \coq{InOp} is generalised to a \emph{recursive
coalgebra}~\cite{AdamekMM19,CaprettaUV04}, that is intuitively a coalgebra that
can be applied only a finite amount of times.  In fact, in the case
of \coq{divide} this is a recursive coalgebra because the size of the list is
getting smaller.

@@

  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
This definition, however, prevents us from composing catamorphisms and
anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular:
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite
trees, then their return type can also be the \emph{least} fixed point of the
container extension. We encode in Coq the notion of \emph{recursive
anamorphisms} and \emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  Recursive
coalgebras are coalgebras whose anamorphism produces finite trees.  The
following predicate captures this fact, \coq{h} is recursive if
\coq{h} can only be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times (\coq{RecF}).  Note
that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
